Digging Deeper Into the State of the Practice for Domain Specific Research Software

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Shortened paper to just over 14 pages.

	We have developed a methodology for assessing the state of the software
	development practices for a given research software domain. Our methodology
	prescribes the following steps: i) Identify the domain; ii) Identify a list
	of candidate software packages; iii) Filter the list to a length of about 30
	packages; iv) Collect repository related data on each software package, like
	number of stars, number of open issues, number of lines of code; v) Fill in
	the measurement template (the template consists of 108 questions to assess 9
	qualities (including the qualities of installability, usability and
	visibility)); vi) Interview developers (the interview consists of 20
	questions and takes about an hour); and, vii) Rank the software using the
	Analytic Hierarchy Process (AHP). The data is analyzed to: i) comparing the
	ranking by best practices against an estimate of the communities ranking of
	popularity; ii) comparing artifacts, tools and processes to current research
	software development guidelines; and, iii) exploring pain points. We
	estimate the time to complete an assessment for a given domain at 173 person
	hours. We illustrate our method via the example of Lattice Boltzmann
	Solvers, where we find that the top packages engaged in most of recommended
	best practices, but show room for improvement with respect to providing API
	documentation, a roadmap, a code of conduct, programming style guide,
	uninstall instructions and continuous integration.

@@

The surveys used in previous studies have tended to recruit participants from
all domains of research software, possibly distinguishing them by programming
language (for instance, R developers \cite{PintoEtAl2018}), or by the role of
the developers (for instance postdoctoral researchers \cite{UditAndKatz2017}).
Case studies, on the other hand, have focused on a few specific examples at a
time, as is the nature of case studies.  For our new methodology, we propose a
scope between these two extremes.  Rather than focus on all research software,
or just a few examples, we will focus on one domain of research software at a
time. The practical reason for defining a domain scope is that digging deep into
repository data takes time, making a broad scope infeasible. We have determined
that a practical constraint for the work load is for a team as small as one
individual doing one person month worth of work per domain.\footnote{A person
month is considered to be $20$ working days ($4$ weeks in a month, with $5$ days
of work per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person
hours.} Focusing on one domain at a time has more than just practical
advantages.  By restricting ourselves to a single domain we can bring domain
knowledge and domain experts into the mix.  The domain customized insight
provided by the assessment has the potential to help a specific domain as they
adopt and develop new development practices.  Moreover, measuring different
domains facilitates uncovering domain specific practices, which can later be
compared and contrasted.

@@

Only so much information can be gleaned by digging into software repositories.
To gain additional insight, we need to interview developers
(Section~\ref{SecSurvey}).  We need to learn their concerns and how they deal
with these concerns; we need to learn their pain points.

@@

Close to our goal of 160 person hours, we estimate 173 hours to complete the
assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes the
domain has been decided, the Domain Expert has been recruited, and the template
spreadsheets and AHP tool~\cite{SmithEtAl2021}, will be utilized. The hours are
a rough estimate, based on our experience completing assessments for medical
image analysis~\cite{Dong2021} and the Lattice Boltzmann Method
(LBM)~\cite{Michalski2021}.  The estimate should be adjusted if the number of
packages measured differs from 30.

@@

A domain of research software must be identified to begin the assessment. To be
applicable for the methodology described in this document, the chosen domain
must have the following properties:

\begin{enumerate}
\item The domain must have well-defined and stable theoretical underpinning.  A
  good sign of this is the existence of standard textbooks, preferably
  understandable by an upper year undergraduate student.
\item There must be a community of people studying the domain.
\item The software packages must have open source options.
\item A preliminary search, or discussion with experts, suggests that there will
  be numerous, close to 30, candidate software packages in the domain that
  qualify as `research software'.
\end{enumerate}

@@

To answer~\rqref{RQ_WhatProjects} we needed to identify existing projects. The
candidate software should be found through search engine queries targeting
authoritative lists of software.  Candidate places to search include
\href{https://github.com/} {GitHub} and \href{https://swmath.org/} {swMATH}, as
well as through scholarly articles. The Domain Expert
(Section~\ref{sec_vet_software_list}) should also be engaged in selecting the
candidate software.  The following properties were considered when creating the
list and reviewing the candidate software:

\begin{enumerate}
	\item The software functionality must fall within the identified domain.
	\item The source code must be viewable.
	\item The repository based measures should be available, which implies a
	preference for GitHub-style repositories.
	\item The software cannot be marked as incomplete.
\end{enumerate}

@@

If the list of software is too long (over around 30 packages)
for~\rqref{RQ_WhatProjects}, then filters are applied in the priority order
listed. Copies of both the initial and filtered lists, along with the rationale
for shortening the list, should be kept for traceability purposes.

@@

For the LBM example the initial list had 46 packages~\cite{Michalski2021}. This
list was filtered by scope, usage, and age to decrease the length to 24. Many of
the removed packages could not be tested as there was no installation guide,
they were incomplete, source code was not publicly available, a license was
needed, or the project was out of scope or not up to a standard that would
support incorporating them into this study~\cite{Michalski2021}.

@@

Once we have measured each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
(AHP), a decision-making technique that is used to compare multiple options by
multiple criteria \cite{Saaty1980}. AHP performs a pairwise analysis using a
matrix and generates an overall score as well as individual quality scores for
each software package. The advantage of pair-wise comparisons is they facilitate
a separation of concerns.  Rather than worry about the entire problem, the
decision maker can focus on one comparison at a time.  In our work AHP performs
a pairwise analysis between each of the 9 quality options for each of the
(approximately) 30 software packages.  This results in a matrix, which is used
to generate an overall score for each software package for the given
criteria~\cite{SmithEtAl2016}.

@@

Figure~\ref{Fig_OverallScore} shows the overall ranking of the LBM software
packages. In the absence of other information on priorities, the overall ranking
was calculated by assuming an equal weighting between all qualities. If the
qualities were weighed differently, the overall software package ranking would
change.

@@

Our methodology relies on engaging a Domain Expert to vet the list of projects
(\rqref{RQ_WhatProjects}) and the AHP ranking (\rqref{RQ_HighestQuality}).  The
Domain Expert is an important member of the assessment
team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
software, or try to definitively rank the software. Non-experts have the problem
that they can only rely on information available on-line, which has the
following drawbacks:
\begin{inparaenum}[i)]
  \item the on-line resources could have false or inaccurate information; and,
  \item the on-line resources could leave out relevant information that is so
in-grained with experts that nobody thinks to explicitly record it.
\end{inparaenum}
Domain experts may be recruited from academia or industry.  The only
requirements are knowledge of the domain and a willingness to be engaged in the
assessment process.  The Domain Expert does not have to be a software developer,
but they should be a user of domain software.  Given that the domain experts are
likely to be busy people, the measurement process cannot put to much of a burden
on their time.

@@

Since each domain we will study will have a reasonably small scope, we will be
able to view the software as constituting a program family. Studying the common
properties within a family of related programs is termed a domain analysis.  For
the current methodology time constraints necessitate a shallow analysis. For
each domain a table should be constructed that distinguishes the programs under
study by their variabilities.  In research software the variabilities are often
related to assumptions.  Table~\ref{tbl_features} shows the variabilities for
the LBM software example~\cite{Michalski2021}.

@@

Table~\ref{repometrics} compares the AHP ranking of the LBM package versus their
popularity in the research community.  Nine packages do not use GitHub, so they
do not have a measure of repository stars. Looking at the repository stars of
the other 15 packages, we can observe a pattern where packages that have been
highly ranked by our assessment tend to have more stars than lower ranked
packages. The best ranked package by AHP (ESPResSo) has the second most stars,
while the ninth ranked package (Sailfish) has the highest number of stars. The
same correlation is observed in the repository watch column, although this
column contains less data, since two of the packages (Palabos, waLBerla) use
GitLab, which does not track watches. Although the AHP ranking and the community
popularity estimate are not perfect measures, they do suggest a correlation
between best practices and popularity.

\begin{table}[!h]
	\begin{center}
		\begin{tabular}{ p{3cm}p{1.25cm}p{1.75cm}p{1.75cm}p{1.75cm}p{1.75cm} }
			\toprule
			Name & Our Ranking & Repository Stars & Repository Star Rank &
			Repository Watches & Repository Watch Rank\\
			\midrule
			ESPResSo & 1 & 145 & 2 & 19& 2\\
			Ludwig & 2 & 27 & 8 & 6& 7\\
			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
			OpenLB & 4 & No Git & No Git & No Git& No Git\\
			LUMA & 5 & 33 & 7 & 12& 4\\
			pyLBM & 6 & 95 & 3 & 10& 5\\
			DL\_MESO (LBE) & 7 & No Git & No Git & No Git & No Git\\
			Musubi & 8 & No Git & No Git & No Git & No Git\\
			Sailfish & 9 & 186 & 1 & 41& 1\\
			waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
			laboetie & 11 & 4 & 13 & 5& 8\\
			TCLB & 12 & 95 & 3 & 16& 3\\
			MechSys & 13 & No Git & No Git & No Git& No Git\\
			lettuce & 14 & 48 & 4 & 5& 8\\
			ESPResSo++ & 15 & 35 & 5 & 12& 4\\
			MP-LABS & 16 & 12 & 11 & 2& 9\\			
			SunlightLB & 17 & No Git & No Git & No Git& No Git\\
			LB3D & 18 & No Git & No Git & No Git& No Git\\			
			LIMBES & 19 & No Git & No Git & No Git& No Git\\
			LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
			HemeLB & 21 & 12 & 11 & 12& 4\\
			lbmpy & 22 &  11 & 12 & 2& 9\\	
			LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
			LatBo.jl & 24 & 17 & 10 & 8& 6\\			
			\bottomrule
		\end{tabular}
		\caption{Repository Ranking Metrics} \label{repometrics}
	\end{center}
	\end{table}

@@

In this step of the methodology we answer \rqref{RQ_CompareArtifacts} by
comparing the artifacts that we observed in the domain repositories to those
observed and recommended for research software in general.  While filling in the
measurement template (Section~\ref{empiricalmeasures}), the domain software is
examined for the presence of artifacts, which are then categorized by frequency.
We suggest grouping them into the categories: common (more than 2/3 of
projects), less common (between 1/3 and 2/3), and rare (less than 1/3 of
projects). The observed frequency of artifacts should then be compared to the
artifacts recommended by research software guidelines, as summarized in
Table~\ref{Tbl_Guidelines}.

@@

Once the data on tools is collected, the use of two specific tools should be
compared to research software norms: version control and continuous integration.
The poor adoption of version control tools that Wilson lamented in 2006
\cite{Wilson2006} has greatly improved in the intervening years.  A little over
10 years ago version control was estimated to be used in only 50\% of research
software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
survey in 2018 shows 81\% of developers use a version control system
\cite{AlNoamanyAndBorghi2018}.  \cite{Smith2018} has similar results, showing
version control usage for alive projects in mesh generation, geographic
information systems and statistical software for psychiatry increasing from
75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
a four year period ending in 2018.  (For completeness the same study showed a
decrease in version control usage for seismology software over the same time
period, from 41\% down to 36\%).  Almost every software guide cited in
Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
For LBM packages 67\% of LBM packages use version control (GitHub, GitLab or
CVS). The high usage of version control tools in LBM software matches the trend
in research software in general.

@@

To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
compare their responses to pain points identified in the
literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
include: lack of development time, Cross-platform compatibility, interruptions
while coding, scope bloat, lack of user feedback, hard to collaborate on
software projects, aloneness, dependency management, data handling concerns
(like data quality, data management and data privacy), reproducibility, and
software scope determination.

An example pain point noted by LBM developers is a lack of development time. A
developer of pyLBM noted that their small development team did not have enough
time to implement new features. Small development teams are common for LBM
software packages (as shown in the measurement table excerpt in
Figure~\ref{measurement_template_image}).

>>>>

	We have developed a methodology for assessing the state of the software
	development practices for a given research software domain. Our methodology
	prescribes the following steps: i) Identify the domain; ii) Identify a list
	of candidate software packages; iii) Filter the list to a length of about 30
	packages; iv) Collect repository related data on each software package, like
	number of stars, number of open issues, number of lines of code; v) Fill in
	the measurement template (the template consists of 108 questions to assess 9
	qualities (including the qualities of installability, usability and
	visibility)); vi) Rank the software using the Analytic Hierarchy Process
	(AHP); vii) Interview developers (the interview consists of 20 questions and
	takes about an hour); and, viii) conduct a domain analysis. The collected
	data is analyzed by: i) comparing the ranking by best practices against an
	the ranking by popularity; ii) comparing artifacts, tools and processes to
	current research software development guidelines; and, iii) exploring pain
	points. We estimate the time to complete an assessment for a given domain at
	173 person hours. The method is illustrated via the example of Lattice
	Boltzmann Solvers, where we find that the top packages engaged in most of
	recommended best practices, but still show room for improvement with respect
	to providing API documentation, a roadmap, a code of conduct, programming
	style guide, uninstall instructions and continuous integration.

@@

The surveys used in previous studies have tended to recruit participants from
all domains of research software, possibly distinguishing them by programming
language (for instance, R developers~\cite{PintoEtAl2018}), or by the role of
the developers (for instance postdoctoral researchers \cite{UditAndKatz2017}).
Case studies, on the other hand, have focused on a few specific examples at a
time, as is the nature of case studies.  For our new methodology, we propose a
scope between these two extremes.  Rather than focus on all research software,
or just a few examples, we will focus on one domain of research software at a
time. The practical reason for this scope is that digging deep into
repository data takes time, making a broad scope infeasible. We have imposed
a practical constraint of one person month worth of work per domain.\footnote{A person
month is considered to be $20$ working days ($4$ weeks in a month, with $5$ days
of work per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person
hours.} Focusing on one domain at a time has more than just practical
advantages.  By restricting ourselves to a single domain we can bring domain
knowledge and domain experts into the mix.  The domain customized insight
provided by the assessment has the potential to help a specific domain as they
adopt and develop new development practices.  Moreover, measuring different
domains facilitates uncovering domain specific practices, which can later be
compared and contrasted.

@@

Only so much information can be gleaned by digging into software repos.
To gain additional insight, we need to interview developers
(Section~\ref{SecSurvey}) to learn:

@@

Close to our goal of 160 person hours, we estimate 173 hours to complete the
assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes the
domain has been decided, the Domain Expert has been recruited, and the template
spreadsheets and AHP tool~\cite{SmithEtAl2021}, will be utilized.

@@

To be applicable for the methodology described in this document, the chosen
domain must have the following properties:

\begin{enumerate}
\item The domain must have well-defined and stable theoretical underpinning.
\item There must be a community of people studying the domain.
\item The software packages must have open source options.
\item A preliminary search, or discussion with experts, suggests that there will
  be around 30 or more candidate software packages.
\end{enumerate}

@@

The candidate software to answer~\rqref{RQ_WhatProjects} should be found through
search engine queries,  
\href{https://github.com/} {GitHub}, \href{https://swmath.org/} {swMATH}
and scholarly articles. The Domain Expert (Section~\ref{sec_vet_software_list})
should also be engaged in selecting the candidate software.  The following
properties are considered when creating the list:

\begin{enumerate}
	\item The software functionality must fall within the identified domain.
	\item The source code must be viewable.
	\item The repository based measures should be available.
	\item The software must be complete.
\end{enumerate}

@@

If the list of software is too long (over around 30 packages), then filters are
applied in the priority order listed. Copies of both lists, along with the
rationale for shortening the list, should be kept for traceability purposes.

@@

For the LBM example the initial list had 46 packages~\cite{Michalski2021}. This
list was filtered by scope, usage, and age to decrease the length to 24.

@@

Once we have measured each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
(AHP), a decision-making technique that uses pair-wise comparisons used to
compare multiple options by multiple criteria \cite{Saaty1980}. In our work AHP
performs a pairwise analysis between each of the 9 quality options for each of
the (approximately) 30 software packages.  This results in a matrix, which is
used to generate an overall score for each software package for the given
criteria~\cite{SmithEtAl2016}.

@@

Figure~\ref{Fig_OverallScore} shows the overall ranking of the LBM software
packages. In the absence of other information on priorities, the overall ranking
was calculated by assuming an equal weighting between all qualities.

@@

Our methodology relies on engaging a Domain Expert to vet the list of projects
(\rqref{RQ_WhatProjects}) and the AHP ranking (\rqref{RQ_HighestQuality}).  The
Domain Expert is an important member of the assessment
team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
software, or try to definitively rank the software. Non-experts have the problem
that they can only rely on information available on-line, which has the
following drawbacks:
\begin{inparaenum}[i)]
  \item the on-line resources could have false or inaccurate information; and,
  \item the on-line resources could leave out relevant information that is so
in-grained with experts that nobody thinks to explicitly record it.
\end{inparaenum}
Domain experts may be recruited from academia or industry.  The only
requirements are knowledge of the domain and a willingness to be involved.

@@

Since each domain will have a reasonably small scope, we will be able to view
the software as a program family. Studying the common properties within a family
via a domain analysis.  For the current methodology time constraints necessitate
a shallow analysis. For each domain a table should be constructed that
distinguishes the programs under study by their variabilities.  In research
software the variabilities are often related to assumptions.
Table~\ref{tbl_features} shows the variabilities for the LBM
software~\cite{Michalski2021}.

@@

Table~\ref{repometrics} compares the AHP ranking of the LBM package to their
popularity in the research community.  Nine packages do not use GitHub, so they
do not have a measure of repository stars. Looking at the repository stars of
the other 15 packages, we can observe a pattern where packages that have been
highly ranked by our assessment tend to have more stars than lower ranked
packages. The best ranked package by AHP (ESPResSo) has the second most stars,
while the ninth ranked package (Sailfish) has the highest number of stars.
Although the AHP ranking and the community popularity estimate are not perfect
measures, they do suggest a correlation between best practices and popularity.

\begin{table}[!h]
	\begin{center}
		\begin{tabular}{ p{3cm}p{1.25cm}p{1.75cm}p{1.75cm}p{1.75cm}p{1.75cm} }
			\toprule
			Name & Our Ranking & Repository Stars & Repository Star Rank &
			Repository Watches & Repository Watch Rank\\
			\midrule
			ESPResSo & 1 & 145 & 2 & 19& 2\\
			Ludwig & 2 & 27 & 8 & 6& 7\\
			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
			OpenLB & 4 & No Git & No Git & No Git& No Git\\
			LUMA & 5 & 33 & 7 & 12& 4\\
			pyLBM & 6 & 95 & 3 & 10& 5\\
			DL\_MESO (LBE) & 7 & No Git & No Git & No Git & No Git\\
			Musubi & 8 & No Git & No Git & No Git & No Git\\
			Sailfish & 9 & 186 & 1 & 41& 1\\
			waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
			laboetie & 11 & 4 & 13 & 5& 8\\
			TCLB & 12 & 95 & 3 & 16& 3\\
			% MechSys & 13 & No Git & No Git & No Git& No Git\\
			% lettuce & 14 & 48 & 4 & 5& 8\\
			% ESPResSo++ & 15 & 35 & 5 & 12& 4\\
			% MP-LABS & 16 & 12 & 11 & 2& 9\\			
			% SunlightLB & 17 & No Git & No Git & No Git& No Git\\
			% LB3D & 18 & No Git & No Git & No Git& No Git\\			
			% LIMBES & 19 & No Git & No Git & No Git& No Git\\
			% LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
			% HemeLB & 21 & 12 & 11 & 12& 4\\
			% lbmpy & 22 &  11 & 12 & 2& 9\\	
			... & ... & ... & ... & ... & ...\\	
			LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
			LatBo.jl & 24 & 17 & 10 & 8& 6\\			
			\bottomrule
		\end{tabular}
		\caption{Excerpt from Repository Ranking Metrics~\cite{Michalski2021}}
		\label{repometrics}
	\end{center}
	\end{table}

@@

We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that we observe
to those observed and recommended for research software in general.  While
filling in the measurement template (Section~\ref{empiricalmeasures}), the
domain software is examined for the presence of artifacts, which are then
categorized by frequency as: common (more than 2/3 of projects), less common
(between 1/3 and 2/3), and rare (less than 1/3). The observed frequency of
artifacts should then be compared to the artifacts recommended by research
software guidelines, as summarized in Table~\ref{Tbl_Guidelines}.

@@

Once the data on tools is collected, the use of two specific tools should be
compared to research software norms: version control and continuous integration.
A little over 10 years ago version control was estimated to be used in only 50\%
of research software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. More
recently, version control usage rates for active mesh generation, geographic
information system and statistical software packages were close to
100~\cite{Smith2018}.  Almost every software guide cited in
Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
For LBM packages 67\% of LBM packages use version control (GitHub, GitLab or
CVS). The high usage of version control tools in LBM software matches the trend
in research software in general.

@@

To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
then compare their responses to pain points identified in the
literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
include: lack of development time, cross-platform compatibility, scope bloat,
lack of user feedback, dependency management, data handling concerns,
reproducibility, and software scope determination.

An example pain point noted by LBM developers is a lack of development time. A
developer of pyLBM noted that their small development team did not have enough
time to implement new features. Small development teams are common for LBM
software packages (as shown in Figure~\ref{measurement_template_image}).
